---
title: "Biost544_hw4"
author: Ziyuan Wang
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(ggplot2)
library(dplyr)
swan <- read_dta("~/Downloads/SWANdata_visit0/28762-0001-Data.dta")
load("/Users/marcy/Downloads/ICPSR_32961/DS0001/32961-0001-Data.rda")
swan10 <- da32961.0001
```
# Processing the dataset
I want to try the simple way of analysis first, that is to look at baseline data and compare bone mineral density in patients of different smoking status, while adjusting for covariates as confounders and effect modifers. From the swan baseline dataset, I extract the variables SWANID (ID), SPBMDT0 (total spine BMD), AGE0 (age at baseline), HEIGHT0 (height at baseline), WEIGHT0 (weight at baseline), BMI0 (BMI at baseline), DTVITD0 (VD supplement), SMOKERE0 (ever smoked regularly).
```{r}
# subset the variables of interest from the swan dataset
swan.keep <- swan[,c("SWANID","SPBMDT0", "AGE0","HEIGHT0","WEIGHT0",
                     "BMI0","DTVITD0","SMOKERE0","HHMEMSM0", "INCOME0")]
# recode the missing values to NA
swan.keep[swan.keep == -1] <- NA
swan.keep[swan.keep == -8] <- NA
swan.keep[swan.keep == -7] <- NA
swan.keep[swan.keep == -9] <- NA
# assume missingness at random and only use the complete cases
swan.all <- swan.keep[complete.cases(swan.keep), ]
swan.all$SMOKERE0 <- ifelse(swan.all$SMOKERE0 == 2, 1, 0)
```
# Calculate the propensity
Since the trial is observational, we want to account for the effect of variables that can be confounders. We will use the IPW method first and calculate propensities.
```{r}
# get the propensities
propen.model <- glm(as.factor(SMOKERE0) ~ AGE0 + HEIGHT0 + WEIGHT0 + 
                      BMI0 + DTVITD0  + as.factor(HHMEMSM0) + as.factor(INCOME0), 
                    family=binomial, data=swan.all)
propensities <- predict(propen.model, data=swan.all, type="response")
# plot the propensities for the two groups
ggplot(data.frame(propensities=propensities, smoking=as.factor(swan.all$SMOKERE0)), 
       aes(x = propensities, y = ..density.., color = smoking)) + geom_density() 
# alot of propensity close to 1, need truncation
# get the range of propensities
range(propensities)
# truncate the propensities score, don't want close to 1 scores
trunc.propen <- propensities %>% pmin(0.95) %>% pmax(0.05)
```
# Calculate the ratio of the probabiliteis and the differences
Using the propensity scores, we can get the estimate difference by inversing weight by these probabilities. The estimated difference we get is -0.0047. This is a negative value and small difference. Next we want to use the boostrap method to get the corresponding confidence interval.
```{r}
n <- nrow(swan.all)
weights <- rep(0, n)
## for patients who smoke:
representative.propen <- sum(swan.all$SMOKERE0 == 1) / n
actual.propen <- trunc.propen
## get the indices for patients who smoked
smoke.ind <- which(swan.all$SMOKERE0 == 1)
## calculate the weight using IPW for smokers and non-smokers
weights[smoke.ind] <- representative.propen/actual.propen[smoke.ind]
weights[-smoke.ind]<- (1 - representative.propen)/(1- actual.propen[-smoke.ind])
## plot the weights
ggplot(data.frame(weights = weights), aes(x=weights, y =..density..)) + 
  geom_histogram()
## calculate the BMD value for smokers and non-smokers
smoking.BMD.est <- with(swan.all,
                         mean((weights*SPBMDT0)[smoke.ind]))
nonsmoking.BMD.est <- with(swan.all,
                             mean((weights*SPBMDT0)[-smoke.ind]))
## calculate the differences in BMD for the two groups
(diff.est <- smoking.BMD.est - nonsmoking.BMD.est)
```
# Calculate the confidence interval using boostrap
The confidence interval obtained is (-0.01386, 0.00698). The CI contains 0, indicating that the difference may not be significant.
```{r}
# function to calculate the weight
calc_weighted_outcome <- function(outcome, label, props){
  weights <- rep(0, length(outcome))
  
  representative.propen <- mean(label)
  actual.propen <- props

  treat.ind <- which(label == 1)
  weights[treat.ind] <- representative.propen/actual.propen[treat.ind]
  weights[-treat.ind]<- (1 - representative.propen)/(1- actual.propen[-treat.ind])
  
  weighted.outcome <- outcome*weights
  
  return(weighted.outcome)
}
# function to get the effect
calc_stat_weighted <- function(weighted.outcome, label){
  return(mean(weighted.outcome[label == 1]) - mean(weighted.outcome[label == 0]))
}


do_one <- function(dat){ # the association is not distorted, confounding effect the same
  resample.inds <- sample(1:nrow(dat), replace = TRUE)
  resample.dat <- dat[resample.inds,]
  # with resampled data
  propen.model <- glm(as.factor(SMOKERE0) ~ AGE0 + HEIGHT0 + WEIGHT0 + 
                        BMI0 + DTVITD0  + as.factor(HHMEMSM0) + as.factor(INCOME0), 
                      family=binomial, data=resample.dat)
  propensities <- predict(propen.model, data = resample.dat, type = "response")
  trunc.prop <- propensities %>% pmax(0.05) %>% pmin(0.95)
  
  weighted.outcome.resamp <- calc_weighted_outcome(resample.dat$SPBMDT0,
                                                   resample.dat$SMOKERE0,
                                                   trunc.prop)
  mean.diff <- calc_stat_weighted(weighted.outcome.resamp, resample.dat$SMOKERE0)
  return(mean.diff)
}

set.seed(2)
boot.dist <- replicate(1e3,
                    do_one(swan.all))
distance.U.L <- quantile(boot.dist, c(0.05,0.95)) - diff.est
(CI <- diff.est - distance.U.L[2:1])
```
# Rerandomization and calculate p-value
Next we can permute and calculate a p-value. After doing permutation testing for 1000 times, we can get a p-value of 0.549. We failed to reject the null hypothesis that there is no association between smoking history and bone mineral density, while adjusting for potential confounders.
```{r}
## one permuation function
one.propen.per <- function(outcome, propen){
  # number of label permuation
  n <- length(outcome)
  # permute the labels
  label <- rbinom(n,1,propen)
  # empty vector to store weights
  weights <- rep(0,n)
  # pi rep calculation
  representative <- mean(label)
  actual <- propen
  # assign indices
  ind.t <- which(label == 1)
  # calculate weights for both groups
  weights[ind.t] <- (representative/actual)[ind.t]
  weights[-ind.t] <- ((1-representative)/(1-actual))[-ind.t]
  # return the differences of outcome
  return(mean((weights*outcome)[ind.t]) - mean((weights*outcome)[-ind.t]))
}
set.seed(2)
# replicate the differences
rerandomized.diffs <- 
  replicate(1e3, one.propen.per(swan.all$SPBMDT0, trunc.propen))
# plot the rerandomized differences
ggplot(data.frame(diffs = rerandomized.diffs), aes(x = diffs, y = ..density..)) +
  geom_density() + 
  geom_vline(xintercept = diff.est, color = "red")
# calculate p-val
mean(rerandomized.diffs > diff.est)
```
# Using standardization method
The second method we can use is the standardization method using logistical regression. The point estimate of the difference in BMD between the two groups is -0.00351, a little smaller than the point estimate we obtained using the IPW method. 
```{r}
# get the outcome regression with outcome as the BMD value and predictors as the smoking status + confoudners
outcome.regression <- glm(SPBMDT0 ~ as.factor(SMOKERE0) + AGE0 + HEIGHT0 + 
                            WEIGHT0 + BMI0 + DTVITD0  + as.factor(HHMEMSM0) + as.factor(INCOME0), 
                          family = gaussian, data=swan.all)
# probabilities if they had smoked
BMD.smoker <- swan.all %>% mutate(SMOKERE0 = 1)
# probabilities if they had not smoked
BMD.nonsmoker <- swan.all %>% mutate(SMOKERE0 = 0)
# calculate the difference
(standardized.est <- mean(predict(outcome.regression, 
                                   BMD.smoker, 
                                   type = "response") -
                             predict(outcome.regression, 
                                     BMD.nonsmoker, 
                                     type = "response")))


```
# Calculate the confidence interval
The confidence interval obtained is (-0.01157, 0.00524). The CI contains 0, indicating that the difference may not be significant.
```{r}
do_one_std <- function(dat){ # the association is not distorted, confounding effect the same
  resample.inds <- sample(1:nrow(dat), replace = TRUE) 
  resample.dat <- dat[resample.inds,]
  # with resampled data
 outcome.regression <- glm(SPBMDT0 ~ as.factor(SMOKERE0) + AGE0 + HEIGHT0 + 
                            WEIGHT0 + BMI0 + DTVITD0  + as.factor(HHMEMSM0) + as.factor(INCOME0), 
                          family = gaussian, data=resample.dat)
# probabilities if they had smoked
BMD.smoker <-  resample.dat %>% mutate(SMOKERE0 = 1)
# probabilities if they had not smoked
BMD.nonsmoker <- resample.dat %>% mutate(SMOKERE0 = 0)
# calculate the difference
standardized.est <- mean(predict(outcome.regression, 
                                   BMD.smoker, 
                                   type = "response") -
                             predict(outcome.regression, 
                                     BMD.nonsmoker, 
                                     type = "response"))
return(standardized.est)
}

set.seed(2)
boot.dist <- replicate(1e3,
                    do_one_std(swan.all))
distance.U.L <- quantile(boot.dist, c(0.05,0.95)) - standardized.est
(CI <- standardized.est - distance.U.L[2:1])


```
# Calculate p-value for the standardization test
```{r}
# one permuation for standardization test
one.std.per <- function(data){
  perm <- sample(1:nrow(data), replace = FALSE) ## generating shuffled ids
  perm.data <- data
  perm.data$SMOKERE0 = data$SMOKERE0[perm] # mixing everything randomly
  outcome.regression <- glm(SPBMDT0 ~ as.factor(SMOKERE0) + AGE0 + HEIGHT0 + WEIGHT0 + 
                              BMI0 + DTVITD0  + as.factor(HHMEMSM0) + as.factor(INCOME0), 
                            family = gaussian, data=perm.data)
# probabilities if they had smoked
BMD.smoker.perm <- perm.data %>% mutate(SMOKERE0 = 1)
# probabilities if they had not smoked
BMD.nonsmoker.perm <- perm.data %>% mutate(SMOKERE0 = 0)
# calculate the difference
standardized.est <- mean(predict(outcome.regression, 
                                   BMD.smoker.perm, 
                                   type = "response") -
                             predict(outcome.regression, 
                                     BMD.nonsmoker.perm, 
                                     type = "response"))
standardized.est
}
set.seed(2)
# replicate the differences
rerandomized.diffs.std <- 
  replicate(1e3, one.std.per(swan.all))
# plot the rerandomized differences
ggplot(data.frame(diffs = rerandomized.diffs.std), aes(x = diffs, y = ..density..)) +
  geom_density() + 
  geom_vline(xintercept = standardized.est, color = "red")
# calculate p-val
mean(rerandomized.diffs.std > standardized.est)
```